---
title: "Reproducible Research: Peer Assessment 1"
author: "Scott Loseke"
date: "March 7, 2018"
output: 
  html_document:
    keep_md: true
---
#### load appropriate libraries
```{r message=FALSE}
library(dplyr)
library(lattice)
```
##   Loading and preprocessing the data
####  first we set the working directory
```{r}
setwd("C:/rProjects/repResearch/RepData_PeerAssessment1")
```

####  now read in the data
```{r}
dat1 <- read.csv("activity.csv")
```

#### clean up some formats and create some variables for future work 
```{r}
dat1$date <- as.Date(dat1$date, "%Y-%m-%d")
dat1$wday <- weekdays(dat1$date)
dat1$wdwe <- ifelse(dat1$wday %in% c("Saturday", "Sunday"), "weekend", "weekday")
dat1$N <- seq(1, 17568, by=1)
```

##  What is mean total number of steps taken per day?
#### first calculate the total number of steps taken each day to the dataframe dSumstep
#### create a histogram to show the distribution of steps taken each day
#### the average total number of steps taken per day is 10,767, and the median is 10,765.   
```{r}
dSumstep <- as.data.frame(dat1 %>% group_by(date) %>% summarize(sumSteps = sum(steps)))
hist(dSumstep$sumSteps,  nclass=10, main="Frequency Number of Steps Taken per Day", xlab = "Total steps taken each day")
mean(dSumstep$sumSteps, na.rm = TRUE)
median(dSumstep$sumSteps, na.rm = TRUE)
summary(dSumstep$sumSteps, na.rm = TRUE)
```

## What is the average daily activity pattern?
#### there are 288 five minute intervals each day, looks like our individual is active in the
#### morning taking the most steps, 206.2, on average at interval 835 
```{r warning=FALSE}
iAvgstep <- as.data.frame(dat1 %>% group_by(interval) %>% summarize(avgSteps = mean(steps, na.rm = TRUE)))
plot(iAvgstep$interval, iAvgstep$avgSteps, type="l", main = "Average number of steps by time of day",
      xlab = "Time of day, each five minute interval", ylab = "Average steps taken in interval each day")
iAvgstep <- mutate(iAvgstep, cdist = dplyr::cume_dist(iAvgstep$avgSteps) )
iAvgstep[iAvgstep$cdist == 1,]
```

## Imputing missing values
#### the total number of missing values in the dataset is 2,304 of 17,568 observations, or about 13%
```{r}
sum(is.na(dat1$steps))
sum(is.na(dat1$steps))/length(dat1$steps)
```
#### to impute the missing values i'll use the average by weekday or weekend, and the interval to
#### fill missing values, first compute the average by day type and interval
```{r}
iwAvgstep <- as.data.frame(dat1 %>% group_by(interval, wdwe) %>% summarize(avgSteps = mean(steps, na.rm = TRUE)))
head(iwAvgstep)
tail(iwAvgstep)
```
#### now join the average in each observation of the entire data set and find the missing values
```{r}
dat2 <- left_join(dat1, iwAvgstep, by = c("wdwe" = "wdwe", "interval" = "interval") )
dat2$fill <- is.na(dat2$steps)
```
#### now overwrite only the missing values in column 1, the steps with column 7 the average steps
#### and then check a few observations where data changed from missing to existing
```{r}
dat2[dat2$fill==TRUE, 1] <- dat2[dat2$fill==TRUE, 7]
dat2[c(284:293),]
dat2[c(9790:9799),]
```
#### the new average total number of steps taken per day changed slightly to  10,762, and the median is now 10,571.   
```{r}
dSumstep <- as.data.frame(dat2 %>% group_by(date) %>% summarize(sumSteps = sum(steps)))
hist(dSumstep$sumSteps,  nclass=10, main="Frequency Number of Steps Taken per Day", xlab = "Total steps taken each day")
mean(dSumstep$sumSteps, na.rm = TRUE)
median(dSumstep$sumSteps, na.rm = TRUE)
summary(dSumstep$sumSteps, na.rm = TRUE)
```

## Are there differences in activity patterns between weekdays and weekends?
#### when imputing missing values we assumed there would be differences in activity level on the weekends
#### now we'll check this out, we create a new data frame with the average steps by interval and day type
#### our subject moves around less early, but does get more active in the middle of the day and later at night
```{r}
iwAvgstep <- as.data.frame(dat2 %>% group_by(interval, wdwe) %>% summarize(avgSteps = mean(steps, na.rm = TRUE)))
xyplot(avgSteps ~ interval | wdwe, data=iwAvgstep, type="l", layout = c(1, 2), main="Yes, we see different activity level on Weekends")
```